### 2022-09-28 draft.tex
Deploying to gh-pages from @ PoisotLab/ms_metaweb_perspectives@133195b5572b35c48fcc87e35a1dd3894494a8ec 🚀
Deploying to gh-pages from @ PoisotLab/ms_metaweb_perspectives@0c903b9b810520a82beb4ec474a4befef4adb03f 🚀

Treating interactions as probabilistic (as opposed to binary) events is
a more nuanced and realistic way to represent them. Dallas et al. (2017)
suggested that most links in ecological networks are cryptic,
\emph{i.e.} uncommon or hard to observe. This argument echoes Jordano
(2016): sampling ecological interactions is difficult because it
requires first the joint observation of two species, and then the
observation of their interaction. In addition, it is generally expected
weak or rare links to be more prevalent in networks than common or
strong links (Csermely, 2004), compared to strong, persistent links;
this is notably the case in food chains, wherein many weaker links are
key to the stability of a system (Neutel et al., 2002). In the light of
these observations, we expect to see an over-representation of
low-probability (rare) interactions under a model that accurately
predicts interaction probabilities. Yet the original metaweb definition,
and indeed most past uses of metawebs, was based on the presence/absence
of interactions. Moving towards \emph{probabilistic} metawebs, by
representing interactions as Bernoulli events (see \emph{e.g.} Poisot et
al., 2016), offers the opportunity to weigh these rare interactions
appropriately. The inherent plasticity of interactions is important to
capture: there have been documented instances of food webs undergoing
rapid collapse/recovery cycles over short periods of time (\emph{e.g.}
Pedersen et al., 2017). These considerations emphasize why metaweb
predictions should focus on quantitative (preferentially probabilistic)
predictions, and this should constrain the suite of appropriate models
used to predict them.

It is important to recall that a metaweb is intended as a catalogue of
all potential interactions, which is then filtered for a given
application (Morales-Castilla et al., 2015). In a sense, that most
ecological interactions are elusive can call for a slightly different
approach to sampling: once the common interactions are documented, the
effort required in documenting each rare interaction will increase
exponentially. Recent proposals suggest that machine learning algorithms
can also act as data generators (Hoffmann et al., 2019): high quality
observational data can be used to infer core rules underpinning network
structure, and be supplemented with synthetic data coming from
predictive models trained on them, thereby increasing the volume of
information available for analysis. Indeed, Strydom, Catchen, et al.
(2021) suggested that knowing the metaweb may render the prediction of
local networks easier, because it fixes an ``upper bound'' on which
interactions can exist. In this context, a probabilistic metaweb
represents an aggregation of informative priors on the interactions,
elusive information with the potential to boost our predictive ability
(Bartomeus et al., 2016).

@@

\begin{figure}
\hypertarget{fig:illustration}{%
\centering
\includegraphics{figures/illustration.png}
\caption{Illustration of an embedding for an host-parasite metaweb,
using Random Dot Product Graphs. \textbf{A}, decrease in approximation
error as the number of dimensions in the subspaces increases.
\textbf{B}, position of hosts and parasites in the first two dimensions
of their respective subspaces. \textbf{C}, predicted interaction weight
from the RDPG based on the status of the species pair in the metaweb.
\textbf{D}, relationship between the position on the first dimension and
parasite generalism.}\label{fig:illustration}
}
\end{figure}

The embedding of the metaweb holds several pieces of information
(fig.~\ref{fig:illustration}). In panel \textbf{A}, we show that the
\(L_2\) loss (\emph{i.e.} the sum of squared errors) between the
empirical and reconstructed metaweb decreases when the number of
dimensions (rank) of the subspace increases, with an inflection point
around 25 dimensions. As discussed by Runghen et al. (2021), there is
often a trade-off between the number of dimensions to use (more
dimensions are more computationally demanding) and the quality of the
representation. In this instance, accepting \(L_2 = 500\) as an
approximation of the network means that the error for every position in
the metaweb is \(\approx \left(500/(206\times 121)\right)^{1/2}\). In
fig.~\ref{fig:illustration}, panel \textbf{B}, we show the positions of
hosts and parasites on the first two dimensions of the left and right
subspaces. Note that these values largely skew negative, because the
first dimensions capture the coarse structure of the network: most pairs
of species do not interact, and therefore have negative values. In
fig.~\ref{fig:illustration}, panel \textbf{C}, we show the predicted
weight (\emph{i.e.} the result of the multiplication of the RDGP
subspaces at a rank of 25) as a function of whether the interactions are
observed, not-observed, or unknown due to lack of co-occurrence. This
reveals that the observed interactions have higher predicted weights,
although there is some overlap; the usual approach to identify potential
interactions based on this information would be a thresholding analysis,
which is outside the scope of this manuscript (and is done in the papers
cited in this illustration). Note that the values are not bound to the
unit interval, which emphasizes the need for either scaling or clamping
(although thresholding analyses are insensitive to this choice).
Finally, in fig.~\ref{fig:illustration}, panel \textbf{D}, we show that
the embedding, as it captures structural information about the network,
holds ecological information; indeed, the position of the parasite on
the first dimension of the left sub-space is a linear predictor of its
number of hosts.

@@

@@

@@

>>>>

Treating interactions as probabilistic (as opposed to binary) events is
a more nuanced and realistic way to represent them. Dallas et al. (2017)
suggested that most links in ecological networks are cryptic,
\emph{i.e.} uncommon or hard to observe. This argument echoes Jordano
(2016): sampling ecological interactions is difficult because it
requires first the joint observation of two species, and then the
observation of their interaction. In addition, it is generally expected
weak or rare links to be more prevalent in networks than common or
strong links (Csermely, 2004), compared to strong, persistent links;
this is notably the case in food chains, wherein many weaker links are
key to the stability of a system (Neutel et al., 2002). In the light of
these observations, we expect to see an over-representation of
low-probability (rare) interactions under a model that accurately
predicts interaction probabilities.

Yet the original metaweb definition, and indeed most past uses of
metawebs, was based on the presence/absence of interactions. Moving
towards \emph{probabilistic} metawebs, by representing interactions as
Bernoulli events (see \emph{e.g.} Poisot et al., 2016), offers the
opportunity to weigh these rare interactions appropriately. The inherent
plasticity of interactions is important to capture: there have been
documented instances of food webs undergoing rapid collapse/recovery
cycles over short periods of time (\emph{e.g.} Pedersen et al., 2017).
Furthermore, because the structure of the metaweb cannot be known in
advance, it is important to rely on predictive tools that do not assume
a specific network topology for link prediction (Gaucher et al., 2021),
but are able to work on generalizations of the network. These
considerations emphasize why metaweb predictions should focus on
quantitative (preferentially probabilistic) predictions, and this should
constrain the suite of appropriate models used to predict them.

It is important to recall that a metaweb is intended as a catalogue of
all potential (feasible) interactions, which is then filtered for a
given application (Morales-Castilla et al., 2015). It is therefore
important to separate the interactions that happen ``almost surely''
(repeated observational data), ``almost never'' (repeated lack of
evidence \emph{or} evidence that the link is forbidden through
\emph{e.g.} trait mis-match), and interactions with a probability that
lays somewhere in between. In a sense, that most ecological interactions
are elusive can call for a slightly different approach to sampling: once
the common interactions are documented, the effort required in
documenting each rare interaction will increase exponentially (Jordano,
2016). Recent proposals in other fields emphasize the idea that machine
learning algorithms can also act as data generators (Hoffmann et al.,
2019): high quality observational data can be used to infer core rules
underpinning network structure, and be supplemented with synthetic data
coming from predictive models trained on them, thereby increasing the
volume of information available for analysis. Indeed, Strydom, Catchen,
et al. (2021) suggested that knowing the metaweb may render the
prediction of local networks easier, because it fixes an ``upper bound''
on which interactions can exist. In this context, a probabilistic
metaweb represents an aggregation of informative priors on the
biological feasibility of interactions, which is usually hard to obtain
yet has possibly the most potential to boost our predictive ability
(Bartomeus et al., 2016).

@@

\begin{figure}
\hypertarget{fig:illustration1}{%
\centering
\includegraphics{figures/illustration-part1.png}
\caption{Validation of an embedding for a host-parasite metaweb, using
Random Dot Product Graphs. \textbf{A}, decrease in approximation error
as the number of dimensions in the subspaces increases. \textbf{B},
increase in cumulative variance explained as the number of ranks
considered increases; in \textbf{A} and \textbf{B}, the dot represents
the point of inflexion in the curve (at rank 39) estimated using the
finite differences method. \textbf{C}, position of hosts and parasites
in the space of latent variables on the first and second dimensions of
their respective subspaces (the results have been clamped to the unit
interval). \textbf{D}, predicted interaction weight from the RDPG based
on the status of the species pair in the
metaweb.}\label{fig:illustration1}
}
\end{figure}

In fig.~\ref{fig:illustration1}, we focus on some statistical checks of
the embedding. In panel \textbf{A}, we show that the averaged \(L_2\)
loss (\emph{i.e.} the sum of squared errors) between the empirical and
reconstructed metaweb decreases when the number of dimensions (rank) of
the subspace increases, with an inflection at 39 dimensions (out of 120
initially) according to the finite differences method. As discussed by
Runghen et al. (2021), there is often a trade-off between the number of
dimensions to use (more dimensions are more computationally demanding)
and the quality of the representation. In panel \textbf{B}, we show the
increase in cumulative variance explained at each rank, and visualize
that using 39 ranks explains about 70\% of the variance in the empirical
metaweb. This is a different information from the \(L_2\) loss (which is
averaged across interactions), as it works on the eigenvalues of the
embedding, and therefore captures higher-level features of the network.
In panel \textbf{C}, we show positions of hosts and parasites on the
first two dimensions of the left and right subspaces. Note that these
values largely skew negative, because the first dimensions capture the
coarse structure of the network: most pairs of species do not interact,
and therefore have negative values. Finally in panel \textbf{D}, we show
the predicted weight (\emph{i.e.} the result of the multiplication of
the RDGP subspaces at a rank of 25) as a function of whether the
interactions are observed, not-observed, or unknown due to lack of
co-occurrence. This reveals that the observed interactions have higher
predicted weights, although there is some overlap; the usual approach to
identify potential interactions based on this information would be a
thresholding analysis, which is outside the scope of this manuscript
(and is done in the papers cited in this illustration). Because the
values returned from RDPG are not bound to the unit interval, we
performed a clamping of the weights to the unit space, showing a
one-inflation in documented interaction, and a zero-inflation in other
species pairs. This last figure crosses from the statistical into the
ecological, by showing that species pairs with no documented
co-occurrence have weights that are not distinguishable from species
pairs with no documented interactions, suggesting that (as befits a
host-parasite model) the ability to interact is a strong predictor of
co-occurrence.

\begin{figure}
\hypertarget{fig:illustration2}{%
\centering
\includegraphics{figures/illustration-part2.png}
\caption{Ecological analysis of an embedding for a host-parasite
metaweb, using Random Dot Product Graphs. \textbf{A}, relationship
between the number of parasites and position along the first axis of the
right-subspace for all hosts, showing that the embedding captures
elements of network structure at the species scale. \textbf{B}, weak
relationship between the body mass of hosts (in grams) and the position
alongside the same dimension. \textbf{C}, weak relationship between
bodymass of hosts and parasite richness. \textbf{D}, distribution of
positions alongside the same axis for hosts grouped by taxonomic
family.}\label{fig:illustration2}
}
\end{figure}

As the results of fig.~\ref{fig:illustration1} show that we can extract
an embedding of the metaweb that captures enough variance to be
relevant, in fig.~\ref{fig:illustration2}, we relate the values of
latent variables for hosts to different ecologically-relevant data. In
panel \textbf{A}, we show that host with a higher value on the first
dimension have fewer parasites. This relates to the body size of hosts
in the \emph{PanTHERIA} database (Jones et al., 2009), as shown in panel
\textbf{B}: interestingly, the position on the first axis is only weakly
correlated to body mass of the host; this matches well establihed
results showing that body size/mass is not always a direct predictor of
parasite richness in terrestrial mammals (Morand \& Poulin, 1998), a
result we observe in panel \textbf{C}. Finally, in panel \textbf{D}, we
can see how different taxonomic families occupy different positions on
the first axis, with \emph{e.g.} Sciuridae being biased towards higher
values. These results show how we can look for ecological informations
in the output of network embeddings, which can further be refined into
the selection of predictors for transfer learning.

@@

\leavevmode\hypertarget{ref-Gaucher2021OutDet}{}%
Gaucher, S., Klopp, O., \& Robin, G. (2021). Outlier detection in
networks with missing links. \emph{Computational Statistics \& Data
Analysis}, \emph{164}, 107308.
\url{https://doi.org/10.1016/j.csda.2021.107308}

@@

\leavevmode\hypertarget{ref-Jones2009PanSpe}{}%
Jones, K. E., Bielby, J., Cardillo, M., Fritz, S. A., O'Dell, J., Orme,
C. D. L., Safi, K., Sechrest, W., Boakes, E. H., Carbone, C., Connolly,
C., Cutts, M. J., Foster, J. K., Grenyer, R., Habib, M., Plaster, C. A.,
Price, S. A., Rigby, E. A., Rist, J., \ldots{} Purvis, A. (2009).
PanTHERIA: A species‐level database of life history, ecology, and
geography of extant and recently extinct mammals: Ecological Archives
E090‐184. \emph{Ecology}, \emph{90}(9), 2648--2648.
\url{https://doi.org/10.1890/08-1494.1}

@@

\leavevmode\hypertarget{ref-Morand1998DenBod}{}%
Morand, S., \& Poulin, R. (1998). Density, body mass and parasite
species richness of terrestrial mammals. \emph{Evolutionary Ecology},
\emph{12}(6), 717--727. \url{https://doi.org/10.1023/A:1006537600093}
